Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-01 tagless/tagless-final.tex
Done with Section 4.2

\hyphenation{meta-language meta-languages Meta-OCaml}
\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}
\usepackage{mdwlist}

@@

\begin{abstract}
We have built a family of tagless embeddings (interpretations) for a
higher-order typed object
language in a typed metalanguage (current Haskell and ML), 
requiring neither dependent
types, nor generalized algebraic data types, nor a tag elimination 
postprocessing step. That has prevously considered an open problem
(i.e. impossible). 
Our family of evaluators contains 
an interpreter, a compiler, a partial
evaluator, a term-size evaluator and a CPS call-by-name transformer.
We take advantage of staging when available.

The main tool is to take functions seriously: we encode the
de Bruijn or higher-order abstract syntax  of object terms 
not by using constructors for
an algebraic data type but rather by invoking cogen functions, which occur as
free variables in the encoded term.  In other words, we eschew initial
algebras for the wilder world of final algebras.
(Or: we expose the co-algebraic structure of the object language).
Our approach highlights the importance of order-2 type constructors
(rank-2 polymorphism), available with ML functors and
Haskell98 constructor classes. This lets us encode object terms with
translucent types: enough abstraction is available to interpret the
terms in various ways, and yet enought type information is available 
to typecheck the terms and statically assure their interpretation will
never get stuck.

@@

It is straightforward to create an algebraic data type, say in OCaml, to
represent object terms such as those in Figure~\ref{fig:object}.
\begin{code}
type var = VZ | VS of var
type exp = V of var | B of bool
         | L of exp | A of exp * exp | ...
\end{code}
We represent each variable using a unary de Bruijn index.
For example, we represent the object term $(\fun{x}x)\True$ as
\begin{code}
let test1 = A (L (V VZ), B true)
\end{code}

@@

let rec eval0 env = function 
| V v       -> lookup env v
| B b       -> b 
| L e       -> fun x -> eval0 (x::env) e
| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}

@@

The usual solution is to introduce a universal type \citep[\S1.3]
{WalidICFP02} containing both booleans and functions.
\begin{code}
type u = UB of bool | UA of (u -> u) | ...
\end{code}
We can then write a typed interpreter
\begin{code}
let rec eval env = function
| V v       -> lookup env v
| B b       -> UB b
| L e       -> UA (fun x -> eval (x::env) e)
| A (e1,e2) -> match eval env e1 with UA f ->
               f (eval env e2)
\end{code}
whose inferred type is |u list -> exp -> u|. Now we can evaluate
our sample |test1| term.
\begin{code}
let test1r = eval [] test1
val test1r : u = UB true 
\end{code}

@@

In sum, by Church\hyp encoding terms using ordinary functions,
we achieve a tagless evaluator
for a typed object language in a metalanguage with a simple type
system \citep{hindley-principal,milner-theory}.  In this \emph{final}
rather than \emph{initial} approach, both kinds of run-time errors
in \S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. Because the new interpreter uses no
universal type or
pattern matching, it never results in a run-time error and is in fact
total.  Because this totality is obvious not just to us but also to the
metalanguage implementation, we avoid the serious performance penalty
\citep{WalidICFP02} of error checking.  See \citep{Gluck-jones-optimality}
for deeper technical reasons which inevitably lead to these performance
penalties.

@@

Our |Symantics| class includes only separate lifting methods |bool| and
|int|, not a parametrically polymorphic lifting method, for good reason:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Thus polymorphic lift cannot be compiled without intensional type
analysis.  To avoid the need for polymorphic lift, we turn to
a technique due to Asai \ccshan{cite?}\jacques{yes}: build a dynamic term
alongside every static term.

@@

When we come to the higher-order fragment of our language, however, we
stumble. Let us consider |lam f|. Here, |f| is a function from a
|P1 a| value to a |P1 b| value. The result of |lam f| must be either a
static value or a dynamic value -- but which? Unlike |add| or |if_|, 
|lam| cannot really examine its operand. The arguments of |add| are 
of type |P1 int|, which we can examine by pattern-matching and so 
determine if they are static or dynamic. The argument of |lam| is a 
function, whose structure cannot be observed. One may only 
apply |f| to a particular static or dynamic value, and receive another 
value in return. Alas, as in the case |lam (\x -> mul x dynval)|, 
applying the same function to different static values may give 
either static or dynamic results (in the example above, applying 
the function to (int 0) gives a static value in return). One
solution is to be conservative and always use the |C| compiler to
compile |f| to a future-stage expression, which we can tag with
E1. But this defeats the purpose of partial evaluation as we may miss
many opportunities for static computations and simplifications, such
as multiplication by zero in the example above. A better solution is to
delay the binding-type-analysis (BTA) until the function |f| is
applied to a particular argument (that is, until the result of |lam f|
appears as the argument of |app|). In other words, we have to find a way
to incorporate |f| as it is into the P1 data structure. We would like
to write that when |t| is a particular type |a->b|, this data type
should be 
\begin{code}
data P1 (a->b) = S1 (R (P1 a -> P1 b)) (C (a->b)) 
               | E1 (C (a->b))
\end{code}
That is, we need a non-parametric data type, something akin to
type-indexed functions and type-indexed types, as used in 
\citet{Oliveira2005:TypeCase}, where it is dubbed the \emph{typecase}
design pattern.

The problem we encountered here is quite akin to the typed CPS
transform. When transforming abstractions, we come across
non-functorialness. The connection with CPS is not an accident, as we
shall see in~\S\ref{S:CPS}.

@@

Two common ways of providing typecase in Haskell is by using either
generalized algebraic data types (GADT) or type-class functions
(implemented via typeclasses with functional dependencies)
\citet{Oliveira2005:TypeCase}.  These
methods are equivalent, and here we show the GADT; please see |incope1.hs|
in the accompanying source code for the latter.

>>>>

\hyphenation{meta-language meta-languages Meta-OCaml}
\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage[override]{cmtt}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}
\usepackage{mdwlist}

@@

\begin{abstract}
We have built a family of tagless interpretations for a
higher-order typed object
language in a typed metalanguage (Haskell and ML), 
requiring neither dependent
types, nor generalized algebraic data types, nor a tag elimination 
postprocessing step. It has been considered an open (impossible)
problem to do so.
Our family of interpreters contains 
an evaluator, a compiler, a partial
evaluator, a term-size evaluator and a CPS call-by-name transformer.
We take advantage of staging when available.

The main tool is to take functions seriously: we encode the
de-Bruijn or higher-order abstract syntax  of object terms 
not by using constructors for
an algebraic data type but rather by invoking cogen functions, which occur as
free variables in the encoded term.  In other words, we eschew initial
algebras for the wilder world of final algebras.
(Or: we expose the co-algebraic structure of the object language).
Our approach highlights the importance of order-2 type constructors
(rank-2 polymorphism), available with ML functors and
Haskell98 constructor classes. This lets us encode object terms with
translucent types: enough abstraction is available to interpret the
terms in various ways, and yet enought type information is available 
to typecheck the terms and statically assure their interpretation will
never get stuck.

@@

It is straightforward to create an algebraic data type, say in OCaml, to
represent object terms such as those in Figure~\ref{fig:object}.
For brevity, we elide treating integers, conditionals, and fixpoint in
this section.
\begin{code}
type var = VZ | VS of var
type exp = V of var | B of bool
         | L of exp | A of exp * exp | ...
\end{code}
We represent each variable using a unary de Bruijn index.
For example, we represent the object term $(\fun{x}x)\True$ as
\begin{code}
let test1 = A (L (V VZ), B true)
\end{code}

@@

let rec eval0 env = function ...
| V v       -> lookup env v
| B b       -> b 
| L e       -> fun x -> eval0 (x::env) e
| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}

@@

The usual solution is to introduce a universal type \citep[\S1.3]
{WalidICFP02} containing both booleans and functions.
\begin{code}
type u = UB of bool | UA of (u -> u) | ...
\end{code}
We can then write a typed interpreter
\begin{code}
let rec eval env = function ...
| V v       -> lookup env v
| B b       -> UB b
| L e       -> UA (fun x -> eval (x::env) e)
| A (e1,e2) -> match eval env e1 with UA f ->
               f (eval env e2)
\end{code}
whose inferred type is |u list -> exp -> u|. Now we can evaluate
our sample |test1| term.
\begin{code}
let test1r = eval [] test1
val test1r : u = UB true 
\end{code}

@@

In sum, by Church\hyp encoding terms using ordinary functions,
we achieve a tagless evaluator
for a typed object language in a metalanguage with a simple type
system \citep{hindley-principal,milner-theory}.  In this \emph{final}
rather than \emph{initial} approach, both kinds of run-time errors
in \S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. Because the new interpreter uses no
universal type or
pattern matching, it never results in a run-time error and is in fact
total.  Because this totality is obvious not just to us but also to the
metalanguage implementation, we avoid the serious performance penalty
\citep{WalidICFP02} of error checking.  \Citet{Gluck-jones-optimality}
explains deeper technical reasons that inevitably lead to these performance
penalties.

@@

Our |Symantics| class includes only separate lifting methods |bool| and
|int|, not a parametrically polymorphic lifting method, for good reason:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Thus polymorphic lift cannot be compiled without intensional type
analysis.  To avoid the need for polymorphic lift, we turn to
a technique due to Asai \ccshan{cite}: build a dynamic term
alongside every static term.

@@

When we come to functions, however, we stumble.  According to our
definition of~|P1|, a partially evaluated object function, such as the
identity $\fun{x}x$ embedded in Haskell as |lam (\x -> x)|\texttt{ ::
}|P1 (a -> a)|, consists of a dynamic part (type |C1 (a -> a)|) and
maybe a static part (type |R1 (a -> a)|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  Even the same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.

To enable these simplifications, we delay binding-time analysis (BTA)
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as it is into the |P1| data structure: applying the type constructor
|P1| to a function type |a -> b| should yield one of
\begin{code}
data P1 (a -> b) = S1 (P1 a -> P1 b)
                 | E1 (C (a->b))

data P1 (a -> b) = P1 (Maybe (P1 a -> P1 b))
                      (C (a -> b))
\end{code}
That is, we need a nonparametric data type, something akin to
type-indexed functions and type-indexed types, which
\citet{oliveira-typecase} dub the \emph{typecase} design pattern.

Thus typed partial evaluation, like typed CPS transformation,
inductively defines a map from source types to target types that
performs case distinction on the source type.  The connection with CPS
is not an accident, as we shall see in~\S\ref{S:CPS}.

@@

Two common ways of providing typecase in Haskell is by using either
GADT or type-class functions
(implemented via typeclasses with functional dependencies)
\citep{oliveira-typecase}.  These
methods are equivalent, and here we show the GADT; please see |incope1.hs|
in the accompanying source code for the latter.
